---
title: "Enzyme_Analysis"
author: "Andrew Lonergan"
date: "10/23/2019"
output: html_document
---

# Extracellular Enzyme Assay Analysis
## Geology 490 Final Project
#### Andrew Lonergan  
    
**Introduction**  
This is an R markdown file showing analysis of research I conducted in Dr. Andrew Steen's biogeochemistry lab. We will use a variety of packages and sets of packages to analyze this data in R. Tidyverse is a set of packages meant for data science. Lubridate helps ease the complications surrounding dates and times in R. Devtools creates a directory structure for your packages.  

```{r, message = FALSE}
# install.packages("package name") first, then do the following:
library(tidyverse)
library(lubridate)
library(devtools)
```

Now that we have our main packages loaded in, we need one more package: lmstats. Lmstats is a function to provide a vector summary output of linear models. 

```{r, message = FALSE}
# Must have done library(devtools) to complete next step:
install_github("adsteen/lmstats")
library(lmstats)
```

Read in your data from wherever it is stored in your drive

```{r}
mydata <- read.csv("/Users/AL/Desktop/geol490/final_project/2019_490_finaldata.csv")
glimpse(mydata)
```
  
**Adjust data to improve time and date**  
