---
title: "Extracellular Enzyme Assay Analysis from Samples in Panama"
author: "Andrew Lonergan"
date: "Fall 2019"
output: html_document
---

  
    
      
### Introduction

This is an R markdown file showing analysis of research I conducted in Dr. Andrew Steen's biogeochemistry lab. This markdown can be used as a reference for extracellular enzyme assay analyses, or in order to reproduce the analysis we did. 

#### Research background  
The goal of this project is to determine how extracellular enzymes are reacting with substrates in the various hot springs in which the samples were collected. We are hoping to understand if enzymes are following Michaelis Menten kinetics. 
Inline equations can be useful for displaying variables (greek or otherwise) like $K_m$ or $\mu$.

Display equations are set off from the rest of the text.
$$
v_o = \frac{V_{max} \times [S]}{K_m + [S]}
$$ 

### Initial steps  
We will use a variety of packages and sets of packages to analyze this data in R. Tidyverse is a set of packages meant for data science. Lubridate helps ease the complications surrounding dates and times in R. Devtools creates a directory structure for your packages. Lmstats provides vector summaries of linear models.


**tidyverse**, **lubridate**, **devtools**, and **lmstats**: Install the packages first, then load them in 

```{r, message = FALSE}
library(tidyverse)
library(lubridate)
library(devtools)
library(lmstats)
```

### Thermal degradation test
A test of substrates and their stability at high temperatures. We want to ensure that these substrates aren't breaking apart on their own, that way when we have live samples, any fluorometer readings we get can represent enzymatic activity.  
Read in your data from wherever it is stored in your drive. You can use glimpse to make sure it looks correct

```{r, results = FALSE, message = FALSE}
thermaldeg_data <- read_csv("/Users/AL/Desktop/geol490/final_project/2018_10_31_thermaldeg_markdown.csv")

glimpse(thermaldeg_data)
```

**Note**  
Some of my code will not show in this document to keep it concise. For example, I have ran some mutations using lubridate to ensure my time measurements are placed on a similar scale (elapsed time) instead of the exact time of day I took the measurements. The full code for this analysis can be found on my github page. 

```{r, echo = FALSE}
thermaldeg_data <- thermaldeg_data %>%
  mutate(time.posix = hms(time)) %>%
  group_by(sample, substrate, replicate, std, temperature) %>%
  mutate(elapsed.time = as.numeric(time - min(time))/ 3600)
```

#### Plot raw thermal degradation data
Ggplot is a great package for data visualization. It is relatively simple to use and creates professional looking plots. 
```{r}
raw_thermal <- ggplot(thermaldeg_data, aes(x=elapsed.time, 
                                           y=RFU, 
                                           colour = substrate, 
                                           shape = as.factor(replicate))) + 
  
  geom_point() +
  geom_smooth(method = "lm", se = TRUE) +
  expand_limits(ymin = 0) + 
  
  facet_grid(substrate~temperature, scales = "free_y")

print(raw_thermal)
```

### Begin substrate data
After testing the stability of our substrates, we are ready to move forward with our enzyme analysis. Read in your data from your live sample fluorometer measurements. 
```{r, message = FALSE}
panamadata <- read.csv("/Users/AL/Desktop/geol490/final_project/2019_490_finaldata.csv")
```
    
**Note**  
Similar to the thermal degradation data, I have made a few adjustments to make this code more readily usable for the rest of my analysis. This code can be found on my github.

```{r, echo = FALSE, message = FALSE}
panamadata <- read_csv("/Users/AL/Desktop/geol490/final_project/2019_490_finaldata.csv") %>%
  mutate(date.time = ymd_hms(paste("2020-01-01", Time))) %>%
  rename(filter.type = `Filter type`, filter.weight = `Filter weight`) %>%
  group_by(Site, Substrate, Cuvette, filter.type, filter.weight) %>%
  mutate(elapsed.hr = as.numeric(date.time -  min(date.time)) / 3600)
```


#### Plot the raw data
After you have loaded in your data and made any neccessary manipulations to the time, names, or any other parts, you can now plot the raw data. We will use ggplot again for this. 

```{r}
raw_panama <- ggplot(panamadata, aes(x = elapsed.hr,
                                     y = RFU,
                                     color = filter.type)) +
  
  geom_point() +
  geom_smooth(aes(group = Cuvette),
              method = "lm",
              se = FALSE) +
  
  facet_wrap(Substrate~Site, scales = "free") +
  
  theme(text = element_text(size = 6))

print(raw_panama)
```

### Begin analysis 
Here we create a function that will create a dataframe comparing RFU to elapsed time. Lm_stats provides us with information such as the slope, intercept, and p value
```{r}
lm_fun <- function(df) {
  df <- as.data.frame(df)
  lm_stats(df, "elapsed.hr", "RFU")
}
```

#### Calculate slope, standard error and p value
```{r}
slopes <- panamadata %>%
  nest() %>%
  mutate(lmstats_df = map(data, lm_fun)) %>%
  mutate(uncal.slope = map_dbl(lmstats_df, function(x) x[1, "slope"]),
         uncal.slope.se = map_dbl(lmstats_df, function(x) x[1, "slope.se"]),
         rsq = map_dbl(lmstats_df, function(x) x[1, "pval"]))
```
  
  
Make a column in the slopes data for which fluorophore goes with which substrate
```{r, warning = FALSE}
fluorophore_lookup <- data.frame(Substrate = unique(panamadata$Substrate), 
                                 fluorophore = c("AMC", "AMC", "MUB", "MUB", "MUB"))
slopes <- slopes %>%
  left_join(fluorophore_lookup, by = "Substrate")
```


### Calibration data 
Our calibration data is how RFU changes with fluorophore concentration. Combining this with our previous substrate data we will be able to see enzymatic activity.

```{r, message = FALSE}
calib <- read_csv("/Users/AL/Desktop/geol490/final_project/2018_11_29_Calibration_Markdown.csv") %>%
  rename(conc.mM = `concentration (mM)`, fluorophore = sample) %>%
  mutate(conc.uM = conc.mM * 1000) 
```

#### Plot calibration data
```{r}
p_calib <- ggplot(calib, aes(x=conc.uM, 
                             y=RFU, 
                             colour = as.factor(temperature))) + 
  
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) + 
  
  facet_wrap(~fluorophore)

print(p_calib) 
```

#### Create another function
This time for our calibration data
```{r}
calib_lm_fun <- function(df) {
  df <- as.data.frame(df)
  lm_stats(df, "conc.uM", "RFU")
}
```

#### Calculate slopes of the calibration curve
And create a column for it in the dataframe
```{r}
calib_slopes <- calib %>%
  group_by(fluorophore) %>%
  nest() %>%
  mutate(lmstats_df = map(data, calib_lm_fun),
         calib.slope = map_dbl(lmstats_df, function(x) x[1, "slope"])) %>%
  select(fluorophore, calib.slope)
```


### Combine datasets
Here we will combine our two dataets, dividing our slopes by the calibration slopes. First we are finding the mean filter slice weight.

```{r, warning = FALSE}
mean.filter.weight = 10 * mean(panamadata$filter.weight)

rates <- slopes %>%
  left_join(calib_slopes, by = "fluorophore") %>%
  mutate(v0.unnorm = uncal.slope / calib.slope,
         v0.unnorm.se = uncal.slope.se / calib.slope,
         v0.norm = uncal.slope * 10 / 0.120,
         v0.norm.se = uncal.slope.se * (mean.filter.weight / filter.weight) / 0.120) 
```

#### Summarise rates by calulating mean rates
```{r}
rates_summ <- rates %>%
  group_by(Site, Substrate, filter.type) %>%
  summarise(mean.v0 = mean(v0.norm, na.rm = TRUE),
            v0.sd = sd(v0.norm, na.rm  = TRUE))
```

### Final plot
Which will show enzymatic activity for live samples and controls for each site and substrate
```{r, warning = FALSE}
p_rates <- ggplot(rates_summ, aes(x=Site, 
                                  y=mean.v0, 
                                  shape = filter.type)) + 
  
  geom_point() + 
  geom_pointrange(aes(ymin = mean.v0 - v0.sd,
                      ymax = mean.v0 + v0.sd)) +
  
  ylab(expression(paste(v[0], ", ", mu, "mol ", L^{-1}, " ", hr^{-1}))) +
  scale_shape_manual(values = c(1, 19), 
                     name = "Treatment") +
 
  facet_wrap(~Substrate, scales = "free_y") +
  
  theme_classic() +
  
  theme(
        strip.text = element_text(size = 10),
        legend.position = c(0.85, 0.15),
        text = element_text(size = 10),
        axis.text.x = element_text(angle = -45, hjust = 0))

print(p_rates)         
```

